---
title: "Chapter 11 Exercises from [Monte Carlo theory, methods, and examples](https://statweb.stanford.edu/~owen/mc/)"
author: "Frederick J. Boehm"
date: "7/13/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> 11.1. Consider the Montr´eal metro walk of $$\S$$ 11.2, for a walker starting at X0 =
Snowdon. a) Plot PSnowdon(Xn = Berri-UQAM) as a function of n for n = 1, . . . , 1000.
Make a second plot just for n = 1, . . . , 10.

We first recall the notation: $$P_{start}(X_n = end)$$ denotes the probability, when starting at 'start', that $X_n$ takes the value "end". 

So, we need to start at Snowdon and determine the probability that $X_n$ is Berri-UQAM. 

He gives us the transition matrix for a single time unit in Expression 11.10 on page 10 of the pdf.

$$P = \left(\begin{array} 
&  0 & 1/2 & 0 & 1/2 & 0 \\
1/2 & 0 & 1/2 & 0 & 0 \\
0 & 1/3 & 0 & 2/3 & 0 \\
1/4 & 0 & 1/2 & 0 & 1/4\\
0 & 0 & 0 & 1/2 & 1/2
\end{array}\right)$$


Let's enter the matrix P into the R session. We'll then look at the row 2, column 4 element of powers of the transition matrix.

```{r}
(Pmat <- matrix(data = c(0, 1/2, 0, 1/2, 0, 
                        1/2, 0, 1/2, 0,0,
                        0, 1/3, 0, 2/3, 0, 
                        1/4, 0, 1/2, 0, 1/4, 
                        0, 0, 0, 1/2, 1/2), 
               nrow = 5, 
               byrow = TRUE))
```


```{r}
foo <- numeric()
bar <- Pmat
foo[1] <- bar[2, 4]
for (i in 2:1000){
  bar  <- bar %*% Pmat
  foo[i] <- bar[2, 4]
}
plot(foo)
plot(foo[1:10])
```


> b) Suppose that the walker changes behavior and does not linger at Longueuil.
Instead, every visit to Longueuil is immediately followed by a visit to BerriUQAM. Write the transition matrix Pe for this version of the walk. With
this change in the Markov chain, repeat part a.

We code in R the matrix Pe:

```{r}
Pe <- Pmat
Pe[5, 5] <- 0; Pe[5, 4] <- 1
Pe
```

```{r}
foo <- numeric()
bar <- Pe
foo[1] <- bar[2, 4]
for (i in 2:1000){
  bar  <- bar %*% Pe
  foo[i] <- bar[2, 4]
}
plot(foo)
plot(foo[1:10])
```




> c) Now suppose that the walker always has probability 1/2 of lingering.
Specifically, let Pe be the transition matrix from part b, let I5 be the
5 by 5 identity matrix, and define P¯ = (Pe + I)/2. Repeat part a using
the transition matrix P¯.

```{r}
Pminus <- (Pe + diag(5)) / 2
foo <- numeric()
bar <- Pminus
foo[1] <- bar[2, 4]
for (i in 2:1000){
  bar  <- bar %*% Pminus
  foo[i] <- bar[2, 4]
}
plot(foo)
plot(foo[1:10])
```





> d) Find π(Berri-UQAM) for the original walk. One way to do this is by
solving the eigenvalue problem πP = π, taking care to get the left eigenvector corresponding to eigenvalue 1. Compare this probability to that
for the other two walks considered here. Comment also on how quickly
the limiting probability is approached in the three walks.

